---
title: "Reproducible Research: Peer Assessment 1 - Andy Tinkham"
output: 
  html_document:
    keep_md: true
---

```{r Options}
options(scipen = 999)
```

This assignment makes use of data from a personal activity monitoring device. 
This device collects data at 5 minute intervals through out the day. The data 
consists of two months of data from an anonymous individual collected during 
the months of October and November, 2012 and include the number of steps taken 
in 5 minute intervals each day.

## Loading and preprocessing the data
First, we need to read in the data (included in this same repository by default.
If for some reason the file is missing, download it from the course web site).

```{r LoadData, echo=TRUE, }
# If this file got separated from the data file, download the data file again
if (!file.exists("./activity.zip")) {
        fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
        download.file(fileUrl, destfile = "activity.zip", method = "curl")
        rm(fileUrl)
}

require(readr)
dataset <- read_csv("./activity.zip")

# Convert the interval to factors to make it easier to compare across days
dataset$interval <- factor(dataset$interval)

```

## What is mean total number of steps taken per day?
From this dataset, we can calculate the total number of steps per day:
```{r StepsPerDay}
byday <- aggregate(dataset$steps, by = list(Date=dataset$date), FUN = sum)
byday
```

And we can make a histogram to show the frequency of steps in 1000-step increments:
```{r StepsPerDayHistogram}

hist(byday$x, 
     breaks = 25, 
     col = "purple", 
     main = "Frequency of daily step totals", 
     xlab = "Steps", 
     ylab = "Number of days")
axis(side = 1, 
     at = seq(0, 25000, 5000), 
     labels = seq(0, 25000, 5000))
```

From this, we can see that the the range goes from a couple days having less than 
1000 steps to a couple days having more than 20,000 steps. The largest number of
days (10) had between 10,000 and 11,000 steps. 

```{r CalculateMeanMedian}
stepsmean <- mean(byday$x, na.rm=TRUE)
stepsmedian <- median(byday$x, na.rm=TRUE)
```

Across the measurement period, the mean number of steps was `r stepsmean` and 
the median number of steps was `r stepsmedian`.

## What is the average daily activity pattern?
Looking across the days of the measurement period, we can get a picture of the
average daily activity pattern.

```{r CreateTimeSeries}
byinterval <- aggregate(dataset$steps, by = list(Interval=dataset$interval), 
                        FUN = mean, na.rm=TRUE)
byinterval
```

We can plot this as a time series plot:
```{r DailyActivityPatternPlot}
plot(byinterval$Interval, byinterval$x, 
     type = "l", 
     main = "Average steps per interval",
     xlab = "Interval (in HHMM format)",
     ylab = "Average steps in measurement period")

```

```{r IntervalWithMaxSteps}
largestInterval <- byinterval[which.max(byinterval$x), ]
```

During the study period, the most steps on average (`r largestInterval$x`) 
occurred in the `r largestInterval$Interval` interval.

## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
